---
title: "Machine Learning Course Project"
output: html_notebook
---

```{r}
library(readr)
library(dplyr)
library(caret)
```

```{r download data}
training_data_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"

training_file_name <- "pml-training.csv"

if(!file.exists(training_file_name))
  download.file(training_data_url, training_file_name)
```

magnet_dumbbell_z, magnet_forearm_y, magnet_forearm_z changed from int to double. Does that mean I should also change the x,y,z variants for those variables. In general, shouldn't just almost all columns be doubles?

```{r read data}
training <- read_csv(training_file_name, na = "#DIV/0!", col_types = cols(
  X1 = col_skip(), # contains just the row number
  user_name = col_character(),
  raw_timestamp_part_1 = col_integer(),
  raw_timestamp_part_2 = col_integer(),
  cvtd_timestamp = col_character(),
  new_window = col_factor(levels = c("yes", "no")),
  num_window = col_integer(),
  roll_belt = col_double(),
  pitch_belt = col_double(),
  yaw_belt = col_double(),
  total_accel_belt = col_integer(),
  kurtosis_roll_belt = col_double(),
  kurtosis_picth_belt = col_double(),
  kurtosis_yaw_belt = col_character(),
  skewness_roll_belt = col_double(),
  skewness_roll_belt.1 = col_double(),
  skewness_yaw_belt = col_character(),
  max_roll_belt = col_character(),
  max_picth_belt = col_character(),
  max_yaw_belt = col_double(),
  min_roll_belt = col_character(),
  min_pitch_belt = col_character(),
  min_yaw_belt = col_double(),
  amplitude_roll_belt = col_character(),
  amplitude_pitch_belt = col_character(),
  amplitude_yaw_belt = col_double(),
  var_total_accel_belt = col_character(),
  avg_roll_belt = col_character(),
  stddev_roll_belt = col_character(),
  var_roll_belt = col_character(),
  avg_pitch_belt = col_character(),
  stddev_pitch_belt = col_character(),
  var_pitch_belt = col_character(),
  avg_yaw_belt = col_character(),
  stddev_yaw_belt = col_character(),
  var_yaw_belt = col_character(),
  gyros_belt_x = col_double(),
  gyros_belt_y = col_double(),
  gyros_belt_z = col_double(),
  accel_belt_x = col_integer(),
  accel_belt_y = col_integer(),
  accel_belt_z = col_integer(),
  magnet_belt_x = col_integer(),
  magnet_belt_y = col_integer(),
  magnet_belt_z = col_integer(),
  roll_arm = col_double(),
  pitch_arm = col_double(),
  yaw_arm = col_double(),
  total_accel_arm = col_integer(),
  var_accel_arm = col_character(),
  avg_roll_arm = col_character(),
  stddev_roll_arm = col_character(),
  var_roll_arm = col_character(),
  avg_pitch_arm = col_character(),
  stddev_pitch_arm = col_character(),
  var_pitch_arm = col_character(),
  avg_yaw_arm = col_character(),
  stddev_yaw_arm = col_character(),
  var_yaw_arm = col_character(),
  gyros_arm_x = col_double(),
  gyros_arm_y = col_double(),
  gyros_arm_z = col_double(),
  accel_arm_x = col_integer(),
  accel_arm_y = col_integer(),
  accel_arm_z = col_integer(),
  magnet_arm_x = col_integer(),
  magnet_arm_y = col_integer(),
  magnet_arm_z = col_integer(),
  kurtosis_roll_arm = col_double(),
  kurtosis_picth_arm = col_double(),
  kurtosis_yaw_arm = col_double(),
  skewness_roll_arm = col_double(),
  skewness_pitch_arm = col_double(),
  skewness_yaw_arm = col_double(),
  max_roll_arm = col_character(),
  max_picth_arm = col_character(),
  max_yaw_arm = col_character(),
  min_roll_arm = col_character(),
  min_pitch_arm = col_character(),
  min_yaw_arm = col_character(),
  amplitude_roll_arm = col_character(),
  amplitude_pitch_arm = col_character(),
  amplitude_yaw_arm = col_character(),
  roll_dumbbell = col_double(),
  pitch_dumbbell = col_double(),
  yaw_dumbbell = col_double(),
  kurtosis_roll_dumbbell = col_double(),
  kurtosis_picth_dumbbell = col_double(),
  kurtosis_yaw_dumbbell = col_character(),
  skewness_roll_dumbbell = col_double(),
  skewness_pitch_dumbbell = col_double(),
  skewness_yaw_dumbbell = col_character(),
  max_roll_dumbbell = col_character(),
  max_picth_dumbbell = col_character(),
  max_yaw_dumbbell = col_double(),
  min_roll_dumbbell = col_character(),
  min_pitch_dumbbell = col_character(),
  min_yaw_dumbbell = col_double(),
  amplitude_roll_dumbbell = col_character(),
  amplitude_pitch_dumbbell = col_character(),
  amplitude_yaw_dumbbell = col_double(),
  total_accel_dumbbell = col_integer(),
  var_accel_dumbbell = col_character(),
  avg_roll_dumbbell = col_character(),
  stddev_roll_dumbbell = col_character(),
  var_roll_dumbbell = col_character(),
  avg_pitch_dumbbell = col_character(),
  stddev_pitch_dumbbell = col_character(),
  var_pitch_dumbbell = col_character(),
  avg_yaw_dumbbell = col_character(),
  stddev_yaw_dumbbell = col_character(),
  var_yaw_dumbbell = col_character(),
  gyros_dumbbell_x = col_double(),
  gyros_dumbbell_y = col_double(),
  gyros_dumbbell_z = col_double(),
  accel_dumbbell_x = col_integer(),
  accel_dumbbell_y = col_integer(),
  accel_dumbbell_z = col_integer(),
  magnet_dumbbell_x = col_integer(),
  magnet_dumbbell_y = col_integer(),
  magnet_dumbbell_z = col_double(),
  roll_forearm = col_double(),
  pitch_forearm = col_double(),
  yaw_forearm = col_double(),
  kurtosis_roll_forearm = col_double(),
  kurtosis_picth_forearm = col_double(),
  kurtosis_yaw_forearm = col_character(),
  skewness_roll_forearm = col_double(),
  skewness_pitch_forearm = col_double(),
  skewness_yaw_forearm = col_character(),
  max_roll_forearm = col_character(),
  max_picth_forearm = col_character(),
  max_yaw_forearm = col_double(),
  min_roll_forearm = col_character(),
  min_pitch_forearm = col_character(),
  min_yaw_forearm = col_double(),
  amplitude_roll_forearm = col_character(),
  amplitude_pitch_forearm = col_character(),
  amplitude_yaw_forearm = col_double(),
  total_accel_forearm = col_integer(),
  var_accel_forearm = col_character(),
  avg_roll_forearm = col_character(),
  stddev_roll_forearm = col_character(),
  var_roll_forearm = col_character(),
  avg_pitch_forearm = col_character(),
  stddev_pitch_forearm = col_character(),
  var_pitch_forearm = col_character(),
  avg_yaw_forearm = col_character(),
  stddev_yaw_forearm = col_character(),
  var_yaw_forearm = col_character(),
  gyros_forearm_x = col_double(),
  gyros_forearm_y = col_double(),
  gyros_forearm_z = col_double(),
  accel_forearm_x = col_integer(),
  accel_forearm_y = col_integer(),
  accel_forearm_z = col_integer(),
  magnet_forearm_x = col_integer(),
  magnet_forearm_y = col_double(),
  magnet_forearm_z = col_double(),
  classe = col_character()
))
```

# Data exploration
We have data on six different subjects.

```{r}
count(training, user_name)
```

As explained [here](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har) subjects were asked to perform an activity (one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl). They were asked to either perform them correctly (classe A) or wrong in a specific way (classes B-E).

```{r}
count(training, classe)
```

There are three columns containing timestamp information. I don't think these are relevant for predicting. I might be able to deduce a duration from them, which might in turn help with predicting. But at this point, I will ignore them and see if I can get good predictions based on the other variables.

Then there is the new_window en num_window variables. I am not exactly sure what they mean, so I will also not use these for predictions

# Model
## RPart
I wanted to start by using machine learning algorithms such as random forests and regularized regression. The training set is pretty big, so I concered about the time it might take to train the models. To reduce training time, I decided to start training with only 3 variables, just to get a feel for the problem and the training times involved. But even using just 3 variables, training times were already longer then several minutes which I didn't like in this stage.

So I switch to rpart, which runs fast. By manual inspection of the data, I see that most column contain lots of junk (lots of NA). So rather than trying to prune all the bad data column, I decide to start with column that seem to contain usefull and relevant data.

I quickly discover roll_belt:yaw_belt and roll_arm:yaw_arm which give me an accuracy score of around 67%, not too bad.

Then I notice that if I try adding more columns, accuracy actually goes down. Now what? Does this have to do with the algorithm?

I have tried adding several groups of columns, all of them just make accuracy go down. I didn't go through all sensible columns, but I am starting to think that 67% is all the accuracy I am going to get out of rpart.

## Random forest
So I would like to switch to random forests, but how will I get acceptable performance? I am going to start with a small sample of the data.

Met 1000 entries heb ik ineens accuracy 99%!? Met 5000 entries zelfs 100%

```{r}
t <- sample_n(training, 1000)

t <- t %>%
  select(classe 
         ,roll_belt:yaw_belt
         ,roll_arm:yaw_arm#,
#         gyros_arm_x:gyros_arm_z#,
#         accel_arm_x:accel_arm_z
#         ,magnet_arm_x:magnet_arm_z
#         ,magnet_dumbbell_x:magnet_dumbbell_z
#         ,roll_dumbbell:yaw_dumbbell
#         ,total_accel_dumbbell
#         ,gyros_dumbbell_x:gyros_dumbbell_z
#          ,accel_dumbbell_x:accel_dumbbell_z
         )

fit <- caret::train(classe ~ .,
             t, 
             method = "rf")

pred <- predict(fit, t)

confusionMatrix(pred, factor(t$classe))
```