---
title: "Machine Learning Course Project"
output: html_notebook
---

```{r}
library(readr)
library(dplyr)
library(caret)
```

```{r download data}
training_data_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"

training_file_name <- "pml-training.csv"

if(!file.exists(training_file_name))
  download.file(training_data_url, training_file_name)
```

```{r read data}
training <- read_csv(training_file_name, na = "#DIV/0!", col_types = cols(
  X1 = col_skip(), # contains just the row number
  user_name = col_character(),
  raw_timestamp_part_1 = col_integer(),
  raw_timestamp_part_2 = col_integer(),
  cvtd_timestamp = col_character(),
  new_window = col_factor(levels = c("yes", "no")),
  num_window = col_integer(),
  roll_belt = col_double(),
  pitch_belt = col_double(),
  yaw_belt = col_double(),
  total_accel_belt = col_integer(),
  kurtosis_roll_belt = col_double(),
  kurtosis_picth_belt = col_double(),
  kurtosis_yaw_belt = col_character(),
  skewness_roll_belt = col_double(),
  skewness_roll_belt.1 = col_double(),
  skewness_yaw_belt = col_character(),
  max_roll_belt = col_character(),
  max_picth_belt = col_character(),
  max_yaw_belt = col_double(),
  min_roll_belt = col_character(),
  min_pitch_belt = col_character(),
  min_yaw_belt = col_double(),
  amplitude_roll_belt = col_character(),
  amplitude_pitch_belt = col_character(),
  amplitude_yaw_belt = col_double(),
  var_total_accel_belt = col_character(),
  avg_roll_belt = col_character(),
  stddev_roll_belt = col_character(),
  var_roll_belt = col_character(),
  avg_pitch_belt = col_character(),
  stddev_pitch_belt = col_character(),
  var_pitch_belt = col_character(),
  avg_yaw_belt = col_character(),
  stddev_yaw_belt = col_character(),
  var_yaw_belt = col_character(),
  gyros_belt_x = col_double(),
  gyros_belt_y = col_double(),
  gyros_belt_z = col_double(),
  accel_belt_x = col_integer(),
  accel_belt_y = col_integer(),
  accel_belt_z = col_integer(),
  magnet_belt_x = col_integer(),
  magnet_belt_y = col_integer(),
  magnet_belt_z = col_integer(),
  roll_arm = col_double(),
  pitch_arm = col_double(),
  yaw_arm = col_double(),
  total_accel_arm = col_integer(),
  var_accel_arm = col_character(),
  avg_roll_arm = col_character(),
  stddev_roll_arm = col_character(),
  var_roll_arm = col_character(),
  avg_pitch_arm = col_character(),
  stddev_pitch_arm = col_character(),
  var_pitch_arm = col_character(),
  avg_yaw_arm = col_character(),
  stddev_yaw_arm = col_character(),
  var_yaw_arm = col_character(),
  gyros_arm_x = col_double(),
  gyros_arm_y = col_double(),
  gyros_arm_z = col_double(),
  accel_arm_x = col_integer(),
  accel_arm_y = col_integer(),
  accel_arm_z = col_integer(),
  magnet_arm_x = col_integer(),
  magnet_arm_y = col_integer(),
  magnet_arm_z = col_integer(),
  kurtosis_roll_arm = col_double(),
  kurtosis_picth_arm = col_double(),
  kurtosis_yaw_arm = col_double(),
  skewness_roll_arm = col_double(),
  skewness_pitch_arm = col_double(),
  skewness_yaw_arm = col_double(),
  max_roll_arm = col_character(),
  max_picth_arm = col_character(),
  max_yaw_arm = col_character(),
  min_roll_arm = col_character(),
  min_pitch_arm = col_character(),
  min_yaw_arm = col_character(),
  amplitude_roll_arm = col_character(),
  amplitude_pitch_arm = col_character(),
  amplitude_yaw_arm = col_character(),
  roll_dumbbell = col_double(),
  pitch_dumbbell = col_double(),
  yaw_dumbbell = col_double(),
  kurtosis_roll_dumbbell = col_double(),
  kurtosis_picth_dumbbell = col_double(),
  kurtosis_yaw_dumbbell = col_character(),
  skewness_roll_dumbbell = col_double(),
  skewness_pitch_dumbbell = col_double(),
  skewness_yaw_dumbbell = col_character(),
  max_roll_dumbbell = col_character(),
  max_picth_dumbbell = col_character(),
  max_yaw_dumbbell = col_double(),
  min_roll_dumbbell = col_character(),
  min_pitch_dumbbell = col_character(),
  min_yaw_dumbbell = col_double(),
  amplitude_roll_dumbbell = col_character(),
  amplitude_pitch_dumbbell = col_character(),
  amplitude_yaw_dumbbell = col_double(),
  total_accel_dumbbell = col_integer(),
  var_accel_dumbbell = col_character(),
  avg_roll_dumbbell = col_character(),
  stddev_roll_dumbbell = col_character(),
  var_roll_dumbbell = col_character(),
  avg_pitch_dumbbell = col_character(),
  stddev_pitch_dumbbell = col_character(),
  var_pitch_dumbbell = col_character(),
  avg_yaw_dumbbell = col_character(),
  stddev_yaw_dumbbell = col_character(),
  var_yaw_dumbbell = col_character(),
  gyros_dumbbell_x = col_double(),
  gyros_dumbbell_y = col_double(),
  gyros_dumbbell_z = col_double(),
  accel_dumbbell_x = col_integer(),
  accel_dumbbell_y = col_integer(),
  accel_dumbbell_z = col_integer(),
  magnet_dumbbell_x = col_integer(),
  magnet_dumbbell_y = col_integer(),
  magnet_dumbbell_z = col_double(),
  roll_forearm = col_double(),
  pitch_forearm = col_double(),
  yaw_forearm = col_double(),
  kurtosis_roll_forearm = col_double(),
  kurtosis_picth_forearm = col_double(),
  kurtosis_yaw_forearm = col_character(),
  skewness_roll_forearm = col_double(),
  skewness_pitch_forearm = col_double(),
  skewness_yaw_forearm = col_character(),
  max_roll_forearm = col_character(),
  max_picth_forearm = col_character(),
  max_yaw_forearm = col_double(),
  min_roll_forearm = col_character(),
  min_pitch_forearm = col_character(),
  min_yaw_forearm = col_double(),
  amplitude_roll_forearm = col_character(),
  amplitude_pitch_forearm = col_character(),
  amplitude_yaw_forearm = col_double(),
  total_accel_forearm = col_integer(),
  var_accel_forearm = col_character(),
  avg_roll_forearm = col_character(),
  stddev_roll_forearm = col_character(),
  var_roll_forearm = col_character(),
  avg_pitch_forearm = col_character(),
  stddev_pitch_forearm = col_character(),
  var_pitch_forearm = col_character(),
  avg_yaw_forearm = col_character(),
  stddev_yaw_forearm = col_character(),
  var_yaw_forearm = col_character(),
  gyros_forearm_x = col_double(),
  gyros_forearm_y = col_double(),
  gyros_forearm_z = col_double(),
  accel_forearm_x = col_integer(),
  accel_forearm_y = col_integer(),
  accel_forearm_z = col_integer(),
  magnet_forearm_x = col_integer(),
  magnet_forearm_y = col_double(),
  magnet_forearm_z = col_double(),
  classe = col_character()
))
```

# Data exploration
We have data on six different subjects.

```{r}
count(training, user_name)
```

As explained [here](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har) subjects were asked to perform an activity (one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl). They were asked to either perform them correctly (classe A) or wrong in a specific way (classes B-E).

```{r}
count(training, classe)
```

There are three columns containing timestamp information. I don't think these are relevant for predicting. I might be able to deduce a duration from them, which might in turn help with predicting. But at this point, I will ignore them and see if I can get good predictions based on the other variables.

Then there is the new_window en num_window variables. I am not exactly sure what they mean, so I will also not use these for predictions

# Model
## RPart
I wanted to start by using machine learning algorithms such as random forests and regularized regression. The training set is pretty big, so I concered about the time it might take to train the models. To reduce training time, I decided to start training with only 3 variables, just to get a feel for the problem and the training times involved. But even using just 3 variables, training times were already longer then several minutes which I didn't like in this stage.

So I switched to rpart, which is not so sophisticated but runs fast. By manual inspection of the data, I see that a lot of columns contain lots of junk (lots of NA or the same value over and over again). So rather than trying to prune all the bad data columns, I decide to start with columns that seem to contain usefull and relevant data.

I quickly discover that roll_belt through yaw_belt and roll_arm through yaw_arm columns give me an accuracy score of around 67%. This is not too bad given the basic algorithm and the limited number of variables.

Then I notice that if I try adding more columns this doesn´t help and sometimes even lowers accuracy actually. 

I have tried this with several (groups of) variables:

* gyros_arm_x through gyros_arm_z
* accel_arm_x through accel_arm_z
* magnet_arm_x through magnet_arm_z
* magnet_dumbbell_x through magnet_dumbbell_z
* roll_dumbbell through yaw_dumbbell
* total_accel_dumbbell
* gyros_dumbbell_x through gyros_dumbbell_z
* accel_dumbbell_x through accel_dumbbell_z

This is just playing around a bit, nothing rigourous, but at this point I am starting to think that 67% is all the accuracy I am going to get out of rpart.

## Random forest
So I would like to switch to random forests, but how will I get acceptable performance? I don't want to be waiting for 50 minutes, only to discover that I am not using the proper variables. So I decided tos start with a small sample of the data.

Using only 1000 rows from the training set, and only using the variables that worked well with rpart I got a model accuracy of 99%. This is a very nice result and only took about 2 minutes to run.

So I decided to train the model on the full dataset.

```{r}
t <- training %>% select(classe, roll_belt:yaw_belt, roll_arm:yaw_arm)

fit <- caret::train(classe ~ ., t, method = "rf")

pred <- predict(fit, t)

confusionMatrix(pred, factor(t$classe))
```

# Model performance
Evaluate model performance on test set.
